A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2018-12-18 article/04_Discussion.Rmd
language and grammer

Figure \ref{development_burial_type} and \ref{development_burial_construction} show the region wise timeseries of trait occurrence. The figure part A with absolute numbers indicates areas and timeperiods with a lot, and others with very few points of data. For some there is not enough information to make any decision about the relative dominance and for many the general relevance and accuracy of the resulting timeseries can be questioned. There's a variety of reasons for this:

1. The data in Radon-B depends on the projects and interests of the involved researchers: The data was collected as needed for individual regions rather then systematically for all of europe.

2. The availability of ^14^C dates for a region and period in the literature depends on whether a modern research project took place within this context. Besides the research interets of researchers and institutions, there are also political reasons why an area might be better or worse explored.

3. In the field of investigation of this work different research traditions meet, which in the past considered ^14^C data more or less essential for the construction of an absolute chronological assessment. In various Metal Age contexts, 14C data are inferior to the accuracy of relative chronological classifications - e.g. on the basis of primer typology - and are therefore used only sporadically to create absolute chronological fix points. Dendrochronological dating is preferred in moist soil contexts and when the necessary woods are available. The available ^14^C calibration curves form plateaus in some periods, which significantly limit the achievable dating accuracy of the data. If such an effect is known, then no data is ordered at all.

4. ^14^C dating requires a small but sufficiently contextualizable amount of organic finds. Due to the old wood and old water effect, short-lived sample materials such as bones of terrestrial organisms or charred cereal grains are preferred. The majority of ^14^C data for this work comes from bones and teeth. However, their preservation is subject to great differences depending on the soil [@brockPrescreeningTechniquesIdentification2010; @gordonSoilPHBone1981; @nielsen-marshBoneDiagenesisEuropean2007]. In many mineral soil contexts no sufficiently good, dateable material is available. This can affect entire regions, if, for example, poor bone preservation prevails in general as a result of low-calcification soils.

5. From very few regions and periods of the European Bronze Age so many graves are documented that it can be assumed that a significant sample of the burial culture of the entire population has been researched. Instead, it is evident in several contexts that only the burials of individual social groups, gender or ethnic group were recorded: The quantity and nature of the known graves simply cannot be representative of all the deceased. In this case it can be assumed that further, deviant burial rituals were practiced, which left no remains or remains that were difficult to grasp archaeologically. These rituals are not shown in the ^14^C data from Radon-B.

@@

- General and concluding observations

The data representativity and semantic consequences will be discussed below.

In reality, beyond Dunnels distinction, most behaviour is not unambiguously stylistic or functional. For the sake of this case study burial rites are assumed to be stylistic, but in fact their creation causes costs in terms of working time and material to a varying extent. This cost might have been influenced or had their own effect on the human-nature relation: Pyres for example require a significant amount of wood and the erection of barrows several days of manpower. Still it can be expected that these functional disadvantages weren't critical most of the time for the general decision which kind of burial tradition is applied, but maybe for the magnitude of invested energy. Other factors to consider here are the level of grief [@macdonaldGriefBurialAmerican2001] or the representational intentions of the burying relatives [@henrichEvolutionCostlyDisplays2009].

>>>>

The A sections of the figures \ref{development_burial_type} and \ref{development_burial_construction} with absolute grave numbers per region and year indicate areas and periods with a lot and others with very few points of data. For some, there is not enough information to make any decision about the relative dominance and for many, the general relevance and accuracy of the resulting time series can be questioned. There's a variety of reasons for this:

1. The data in Radon-B depends on the projects and interests of the involved researchers: The data was collected as needed for individual regions rather than systematically for all of Europe.

2. The availability of ^14^C dates for a region and period in the literature depends on whether a modern research project took place within this context. Besides the research interests of researchers and institutions, there are also political reasons why an area might be better or worse explored.

3. In the field of investigation of this work different research traditions meet, which in the past considered ^14^C data more or less essential for the construction of an absolute chronological assessment. In various Metal Age contexts, 14C data are inferior to the accuracy of relative chronological classifications - e.g. by primer typology - and are therefore used only sporadically to create absolute chronological fixpoints. Dendrochronological dating is preferred in moist soil contexts and when the necessary woods are available. The available ^14^C calibration curves form plateaus in some periods, which significantly limit the possible dating accuracy of the data. If such an effect is known, then no data is ordered at all.

4. ^14^C dating requires a small but sufficiently contextualizable amount of organic finds. Due to the old wood and old water effect, short-lived sample materials such as bones of terrestrial organisms or charred cereal grains are preferred. The majority of ^14^C data for this work comes from bones and teeth. However, their preservation is subject to huge differences depending on the soil [@brockPrescreeningTechniquesIdentification2010; @gordonSoilPHBone1981; @nielsen-marshBoneDiagenesisEuropean2007]. In many mineral soil contexts no sufficiently good, dateable material is available. This situation can affect entire regions, if, for example, poor bone preservation prevails in general as a result of low-calcification soils.

5. From very few regions and periods of the European Bronze Age so many graves are documented that it can be assumed that a significant sample of the burial culture of the entire population has been researched. Instead, it is evident in several contexts that only the burials of individual social groups, gender or ethnic group were recorded: The quantity and nature of the known graves cannot be representative of all the deceased. In this case, it can be assumed that further, deviant burial rituals were practised, which left no remains or remains that were difficult to grasp archaeologically. These rituals are not shown in the ^14^C data from Radon-B.

Concerning the primary variables *burial type* and *burial construction*, further special effects must be taken into account: The disturbance and robbery of graves was a widespread phenomenon in history and prehistory. Burial mounds were usually attacked more severely than flat graves due to their better visibility and could, therefore, be underrepresented in Radon-B. In many places, they were also the subject of archaeological expeditions in the early days of archaeology at the end of the 19th and beginning of the 20th century. Of course, these excavations did not yield ^14^C-dates. Irrespective of this, it can be assumed that a large part of the burial mounds erected in the Bronze Age was destroyed by agricultural activity. Ploughing could also result in an under-representation of the archaeological findings [@hardingEuropeanSocietiesBronze2000a]. On the other hand, if a burial mound is still present, it is more readily visible and recognised by archaeologists today: Mounds are often clearly identifiable in aerial photographs or digital terrain models, while flat graves are much more difficult to identify for today's remote sensing methods. It is difficult to estimate the impact of these effects in detail, but an overrepresentation of mounds seems more likely than an overrepresentation of flat graves. The dating of cremation graves on bone remains methodically differs from the dating of body graves because it is based on the extraction of the mineral, not the organic, fraction in the bone [@lantingDatingCrematedBones2001]. The underlying chemical method was developed later, has only been used more frequently in the past 15 years and is associated with some particular difficulties [@olsenCharacterisationBlindTesting2008; @strydonck14CDatingCremated2009]. Burnt bone is generally more susceptible to weathering than unburned bone. An under-representation of cremation graves in current data collections is, therefore, more plausible.

@@

In reality, beyond Dunnels distinction, most behaviour is not unambiguously stylistic or functional. For the sake of this case study burial rites are assumed to be stylistic, but in fact, their creation causes costs regarding working time and material to a varying extent. This cost might have been influenced or had their effect on the human-nature relation: Pyres, for example, require a significant amount of wood and the erection of barrows much workforce. Still, it can be expected that these functional disadvantages were not critical most of the time for the general decision which kind of burial tradition is applied, but maybe for the magnitude of invested energy. Other factors to consider here are the level of grief [@macdonaldGriefBurialAmerican2001] or the representational intentions of the burying relatives [@henrichEvolutionCostlyDisplays2009].
